---
title: "preprocessing"
output: html_document
---

# Preprocessing

```{r}
source("http://bioconductor.org/biocLite.R")
library(affy)
library(Biobase)
```

**Question: How to create an AffyBatch object using a data matrix?**
https://www.biostars.org/p/244087/

**Question: Example on constructing AffyBatch object**
https://support.bioconductor.org/p/34261/

**Question: Creating an Expression Set with a csv file**
https://support.bioconductor.org/p/69116/

**Affymetrix CEL files**
http://www.sthda.com/english/wiki/affymetrix-cel-files

**Question: how to convert ExpressionSet to AffyBatch**
https://support.bioconductor.org/p/54812/

**Affybatch Class*
http://web.mit.edu/~r/current/arch/i386_linux26/lib/R/library/affy/html/AffyBatch-class.html

```{r}
?rma
?ExpressionSet
```

# Creating Expression Set with CSV file

```{r}
full_data <- "../../data/genbank/full_data.csv"
full_data <- read.csv(full_data)[,-1]

assay <- full_data[,-c(1:4)]
assay <- as.matrix(assay)
row.names(assay) <- meta[,1]
```

**Expression Set**
https://www.bioconductor.org/packages/release/bioc/vignettes/Biobase/inst/doc/ExpressionSetIntroduction.pdf


```{r}
?AnnotatedDataFrame
meta
```

```{r}
metadata <- data.frame(labelDescription=c("Sample ID number","Sampled tissue","Whether or not patient has endometriosis", "Severity of endometriosis or non-endometriosis"),row.names=c("sample", "tissue", "endometriosis", "severity"))
metadata
```


```{r}
meta  <- full_data[,1:4]
row.names(meta) <- meta[,1]
pheno_data <-new("AnnotatedDataFrame", data=meta, varMetadata=metadata)
# sampleNames(pheno_data)
```

```{r}
sampleNames(assay)
```

```{r}
expression_set <- ExpressionSet(assayData=assay, phenoData=pheno_data)
expression_set
```

^^ Doesn't work.

```{r}
expression_set <- ExpressionSet(assayData=assay, probNames=names(assay))
expression_set
```

**Tutorial**
https://biochem.wisc.edu/sites/default/files/facilities/bcrf/tutorials/R/R_Bioconductor/CH25/from_cel_files_to_gene_list.html#4_preparation

```{r}
try <- rma(expression_set)
```


**GCRMA v. RMA**
https://support.bioconductor.org/p/3248/

**GC Paper**
http://www.biostat.jhsph.edu/~ririzarr/papers/gcpaper.pdf

**Why RMA**
https://support.bioconductor.org/p/63851/

**GCRMA v. Limma**
https://www.biostars.org/p/336470/