---
title: "Draft 1 (Data vis and top diff expression)"
author: "Asem Berkalieva"
date: "11/25/2018"
output: pdf_document
--- 
### Reading in the two datasets 
 
``` {r}
library(ggplot2)
library(dplyr)

# read in PROBESET data
small_subset <- read.csv("../../data/genbank/clean-csv/small_subset.csv")
probeset <- small_subset[,6:105]

# read in GENE data
normal_gene <- read.csv("../../data/figshare/clean-csv/normal.csv")
mild_gene <- read.csv("../../data/figshare/clean-csv/minimal-mild.csv")
severe_gene <- read.csv("../../data/figshare/clean-csv/moderate-severe.csv")

# create a subset
ix <- sample(x=1:ncol(normal_gene), size=100, replace=FALSE)
normal_subset <- normal_gene[,c(ix)]
mild_subset <- mild_gene[,c(ix)]
severe_subset <- severe_gene[,c(ix)]
genes <- rbind(normal_subset, mild_subset, severe_subset)

# check dimensions
dim(probeset)
dim(genes)
```


### Matching genes with probesets: correlation

``` {r}
# obtain correlation for all values between the two datasets
match_genes <- apply(probeset, 1, function(x) {
  which.max(cor(x, t(genes)))
})

# check correlation
cors <- cor(probeset[1:105, 1:100], genes[1:105, 1:100])

# check that the correlation is close to 1
cors <- apply(cbind(probeset, 
                    genes[match_genes, ]), 1, function(x) {
  cor(x[1:100], x[101:200])
})

# check correlation extremes
min(cors)
max(cors)
```

Let's do this one more time to make sure it isn't the random sampling:
``` {r}
# create a new gene subset
ix_new <- sample(x=1:ncol(normal_gene), size=100, replace=FALSE)
normal_subset_2 <- normal_gene[,c(ix_new)]
mild_subset_2 <- mild_gene[,c(ix_new)]
severe_subset_2 <- severe_gene[,c(ix_new)]
genes_v2 <- rbind(normal_subset, mild_subset, severe_subset)

# obtain correlation for all values between the two datasets
match_genes <- apply(probeset, 1, function(x) {
  which.max(cor(x, t(genes_v2)))
})

# check correlation
cors <- cor(probeset[1:105, 1:100], genes_v2[1:105, 1:100])

# check that the correlation is close to 1
cors <- apply(cbind(probeset, 
                    genes_v2[match_genes, ]), 1, function(x) {
  cor(x[1:100], x[101:200])
})

# check correlation extremes
min(cors) 
max(cors)
```


### Exploratory data analysis on GENE dates

This is what the **GENES** look like:
``` {r}
# explore current data of GENES
boxplot(genes,
        main = "Distribution of Genes Expression Data")

# subset
boxplot(genes[,1:5],
        main = "Close up of First Five Genes")
```
 
This is what the **SAMPLES** look like:
``` {r}
# explore current data of SAMPLES
boxplot(t(genes),
        main = "Distribution of Samples")

# subset
boxplot(t(genes)[,1:5],
        main = "Close up of First Five Samples")
```


### Exploratory data analysis on PROBESET dates

This is what the **GENES/columns** look like:
``` {r}
# explore current data of GENES
boxplot(probeset,
        main = "Distribution of Genes Expression Data")

# subset
boxplot(probeset[,1:5],
        main = "Close up of First Five Genes")
```
 
This is what the **SAMPLES/rows** look like:
``` {r}
# explore current data of SAMPLES
boxplot(t(probeset),
        main = "Distribution of Samples")

# subset
boxplot(t(probeset)[,1:5],
        main = "Close up of First Five Samples")
```


### Normalization (work with SMALL SUBSET first)

``` {r}
cut_1 <- read.csv("../../data/genbank/subsets-for-control-removal/cut_1.csv")
```



### Differential expression

``` {r t-test}
# split into two groups
mild <- as.matrix(mild_subset)
severe <- as.matrix(severe_subset)

# run Wilcoxon-rank sum test for each gene
pvals <- c()
for (row in seq(nrow(mild_subset))) {
  p <- wilcox.test(x = mild[row,], severe[row,],
                   exact = FALSE)$p.value
  pvals <- c(pvals, p)
}

# adjust p-values using false-discovery rate method
pvals <- p.adjust(pvals, method = "fdr")

# return top 10 expressianlly dif genes (10 smallest p-values)
gene_pvals <- colnames(mild_subset)
gene_pvals <- data.frame(cbind(gene_pvals, pvals))
gene_pvals_ordered <- gene_pvals[order(gene_pvals$pvals), , drop = FALSE]
```







