---
title: "fgene-code-copy"
author: "Asem Berkalieva"
date: "11/27/2018"
output: pdf_document
---

``` {r}
# load in libraries
library(gcrma)
library(limma)

# read in PROBESET data
small_subset <- read.csv("../../data/genbank/01-small-subset/normalized_subset.csv")
probeset <- small_subset[,6:105]

# work with Cut 1 with controls REMOVED
cut_1_probe <- read.csv("../../data/genbank/02-removed-controls/cut_1.csv")

# create a subset
iv <- sample(x=6:ncol(cut_1_probe), size=100, replace=FALSE)
cut_1 <- cut_1_probe[,c(iv)]

# check dimensions
dim(cut_1)
```

### Exploratory data analysis on PROBESET dates

This is what the **GENES/columns** look like:
``` {r}
# explore current data of GENES
boxplot(probeset,
        main = "Distribution of Genes Expression Data")

# subset
boxplot(probeset[,1:5],
        main = "Close up of First Five Genes")
```

This is what the **SAMPLES/rows** look like:
``` {r}
# explore current data of SAMPLES
boxplot(t(probeset),
        main = "Distribution of Samples")

# subset
boxplot(t(probeset)[,1:5],
        main = "Close up of First Five Samples")
```

\newpage

### Log transform the data

I have log transformed the data to see if it helps... it doesn't.

``` {r log-transform-normalize, fig.align = "center"}
# log transform the data
scaled_log <- log(probeset)

# view the new spread
boxplot(scaled_log,
        main = "Log Transformed Distributions of Probesets")
```

\newpage

### Normalize using LIMMA

Perform basic PCA/EDA on different normalizations. Are we seeing/removing technical variation or biological variation?  

Check assumptions:
- Quantile: assume all have same distribution
- LOESS: check assumptions

Be thorough: justify every step. 

``` {r}
# PROBESET SUBSET
# normalize the data
normalized_quantile <- normalizeBetweenArrays(probeset, method = "quantile")
t_normalized_quantile <- normalizeBetweenArrays(t(probeset), method = "quantile")
normalized_loess <- normalizeBetweenArrays(probeset, method = "cyclicloess")
normalized_scale <- normalizeBetweenArrays(probeset, method = "scale")

# transform to data frame for easy plotting
non_normalized <- as.data.frame(probeset)
normalized_quantile <- as.data.frame(normalized_quantile)
t_normalized_quantile <- as.data.frame(t_normalized_quantile)
normalized_loess <- as.data.frame(normalized_loess)
normalized_scale <- as.data.frame(normalized_scale)

# look at the PROBESETS
par(mfrow=c(1,3))
boxplot(non_normalized,
        main = "Non-normalized")
boxplot(normalized_quantile,
        main = "Normalized Quantile")
boxplot(t(t_normalized_quantile),
        main = "t-Normalized Quantile")
boxplot(normalized_loess,
        main = "Normalized LOESS")
boxplot(normalized_scale,
        main = "Normalized Scale")

# look at the SAMPLES
par(mfrow=c(1,3))
boxplot(t(probeset))
boxplot(t(normalized_quantile),
          main = "Samples with Normalize Quantile")
boxplot(t(normalized_loess),
        main = "Samples with Normalize LOESS")
boxplot(t(normalized_scale),
        main = "Samples with Normalized Scale")
```
 
``` {r}
# CUT 1 
# normalize the data
cut_quantile <- normalizeBetweenArrays(cut_1, method = "quantile")
t_cut_quantile <- normalizeBetweenArrays(t(cut_1), method = "quantile")
cut_loess <- normalizeBetweenArrays(cut_1, method = "cyclicloess")
cut_scale <- normalizeBetweenArrays(cut_1, method = "scale")

# transform to data frame for easy plotting
non_norm_cut <- as.data.frame(cut_1)
cut_quantile <- as.data.frame(cut_quantile)
t_cut_quantile <- as.data.frame(t_cut_quantile)
cut_loess <- as.data.frame(cut_loess)
cut_scale <- as.data.frame(cut_scale)

# look at the PROBESETS
par(mfrow=c(2,3))
boxplot(non_norm_cut,
        main = "Non-normalized")
boxplot(cut_quantile,
        main = "Normalized Quantile")
boxplot(t(t_cut_quantile),
        main = "t-Normalized Quantile")
boxplot(cut_loess,
        main = "Normalized LOESS")
boxplot(cut_scale,
        main = "Normalized Scale")

# look at the SAMPLES
par(mfrow=c(2,3))
boxplot(t(non_norm_cut))
boxplot(t(cut_quantile),
          main = "Samples with Normalize Quantile")
boxplot(t_cut_quantile,
          main = "Samples with t-Normalize Quantile")
boxplot(t(cut_loess),
        main = "Samples with Normalize LOESS")
boxplot(t(cut_scale),
        main = "Samples with Normalized Scale")
```

\newpage

### Normalize using GCRMA

We can't normalize the way that the paper did because they started with .CEL files. We do not have that option. 

``` {r}
expression_set <- ExpressionSet(assayData = as.matrix(probeset))

gcrms <- gcrma(expression_set)

norm <- rma(expression_set)

```


### Working with Figshare/gene data

We want to look at the data from Figshare to see what their "Normalized" data looks like. This way, we can compare and get a standard for what our normalized data should approximately look like.

``` {r}
# read in FIGSHARE/GENE data
normal_gene <- read.csv("../../data/figshare/clean-csv/normal.csv")
mild_gene <- read.csv("../../data/figshare/clean-csv/minimal-mild.csv")
severe_gene <- read.csv("../../data/figshare/clean-csv/moderate-severe.csv")

# create a subset
ix <- sample(x=1:ncol(normal_gene), size=100, replace=FALSE)
normal_subset <- normal_gene[,c(ix)]
mild_subset <- mild_gene[,c(ix)]
severe_subset <- severe_gene[,c(ix)]
genes <- rbind(normal_subset, mild_subset, severe_subset)

# check dimensions
dim(genes)
```

This is what the data looks like for Figshare after they normalized. The distribution for genes are different, but the samples' distributions are similar. 

``` {r}
# look at genes
boxplot(genes)

# look at samples
boxplot(t(genes))
```








