---
title: "fgene-code-copy"
author: "Asem Berkalieva"
date: "11/27/2018"
output: pdf_document
---

``` {r, warnings = FALSE, message = FALSE}
# load in libraries
library(gcrma)
library(limma)
library(dplyr)
```

``` {r, echo = FALSE}
# read in PROBESET data
small_subset <- read.csv("../../data/genbank/01-small-subset/normalized_subset.csv")
probeset <- small_subset[,6:105]

# read in CUT 1 data (controls REMOVED)
cut_1_probe <- read.csv("../../data/genbank/02-removed-controls/cut_1.csv")
 
# create a subset 
set.seed(47)
iv <- sample(x=6:ncol(cut_1_probe), size=100, replace=FALSE)
cut_1 <- cut_1_probe[,c(iv)]

# check dimensions
dim(cut_1)

# read in GCRMA NORMALIZED SUBSET
norm <- read.csv("../../data/genbank/01-normalized/cut_1.csv")

# create a subset 
set.seed(48)
iv <- sample(x=6:ncol(norm), size=100, replace=FALSE)

norm_subset_endo <- norm[,4]
norm_subset_genes <- norm[,6:106]
norm_subset <- cbind(norm_subset_endo, norm_subset_genes)



dim(norm_sub)


```

``` {r, echo = FALSE}
# read in FIGSHARE/GENE data from the final paper
normal_gene <- read.csv("../../data/figshare/clean-csv/normal.csv")
mild_gene <- read.csv("../../data/figshare/clean-csv/minimal-mild.csv")
severe_gene <- read.csv("../../data/figshare/clean-csv/moderate-severe.csv")

# create a subset
set.seed(48)
ix <- sample(x=1:ncol(normal_gene), size=100, replace=FALSE)
normal_subset <- normal_gene[,c(ix)]
mild_subset <- mild_gene[,c(ix)]
severe_subset <- severe_gene[,c(ix)]
genes <- rbind(normal_subset, mild_subset, severe_subset)

# check dimensions
# dim(genes)
```

# RAW PROBESET DATA

### EDA:

This is what the **GENES/columns** look like:
``` {r, echo = FALSE}
# explore current data of GENES
boxplot(probeset,
        main = "Raw Probeset")

# subset
boxplot(probeset[,1:5],
        main = "Close up of First Five Probes")
```

This is what the **SAMPLES/rows** look like:
``` {r, echo = FALSE}
# explore current data of SAMPLES
boxplot(t(probeset),
        main = "Raw Probeset Samples")

# subset
boxplot(t(probeset)[,1:5],
        main = "Close up of First Five Probe Samples")
```

\newpage

### Log transform the data

I have log transformed the raw probeset data to see if it helps... it doesn't.

``` {r}
# log transform the data
scaled_log <- log(probeset)

# view the new spread
boxplot(scaled_log,
        main = "Log Transformed Distributions of Probesets")
```

# CUT 1 DATA

Let's use Cut 1 to see if removing controls now allows a log-transformation to perform better:

``` {r}
# log transform the data
log_cut <- log(cut_1)

# view the new log-transformed data
par(mfrow=c(1,2))

boxplot(cut_1,
        main = "Non-Log Cut 1")
boxplot(log_cut,
        main = "Logged Cut 1")
```

\newpage

# NORMALIZATION

### Normalize using LIMMA

Perform basic PCA/EDA on different normalizations. Are we seeing/removing technical variation or biological variation?  

Check assumptions:
- Quantile: assume all have same distribution
- LOESS: needs to be log-transformed first

**Be thorough: justify every step. **

### RAW PROBESET

``` {r}
# PROBESET SUBSET
# normalize the data
normalized_quantile <- normalizeBetweenArrays(probeset, method = "quantile")
normalized_loess <- normalizeBetweenArrays(probeset, method = "cyclicloess")
normalized_scale <- normalizeBetweenArrays(probeset, method = "scale")

# transform to data frame for easy plotting
non_normalized <- as.data.frame(probeset)
normalized_quantile <- as.data.frame(normalized_quantile)
normalized_loess <- as.data.frame(normalized_loess)
normalized_scale <- as.data.frame(normalized_scale)

# look at the PROBESETS
par(mfrow=c(2,2))
boxplot(non_normalized,
        main = "Non-normalized")
boxplot(normalized_quantile,
        main = "Normalized Quantile")
boxplot(normalized_loess,
        main = "Normalized LOESS")
boxplot(normalized_scale,
        main = "Normalized Scale")

# look at the SAMPLES
par(mfrow=c(2,2))
boxplot(t(probeset))
boxplot(t(normalized_quantile),
          main = "Samples with Normalize Quantile")
boxplot(t(normalized_loess),
        main = "Samples with Normalize LOESS")
boxplot(t(normalized_scale),
        main = "Samples with Normalized Scale")
```
 
### CUT 1: REMOVED CONTROLS
 
``` {r}
# CUT 1 
# normalize the data
cut_quantile <- normalizeBetweenArrays(cut_1, method = "quantile")
cut_loess <- normalizeBetweenArrays(cut_1, method = "cyclicloess")
cut_scale <- normalizeBetweenArrays(cut_1, method = "scale")

# transform to data frame for easy plotting
non_norm_cut <- as.data.frame(cut_1)
cut_quantile <- as.data.frame(cut_quantile)
cut_loess <- as.data.frame(cut_loess)
cut_scale <- as.data.frame(cut_scale)

# look at the PROBESETS
par(mfrow=c(2,2))
boxplot(non_norm_cut,
        main = "Non-normalized")
boxplot(cut_quantile,
        main = "Normalized Quantile")
boxplot(cut_loess,
        main = "Normalized LOESS")
boxplot(cut_scale,
        main = "Normalized Scale")


# look at the SAMPLES
par(mfrow=c(2,2))
boxplot(t(non_norm_cut),
        "Non norm samples")
boxplot(t(cut_quantile),
          main = "Samples with Normalize Quantile")
boxplot(t(cut_loess),
        main = "Samples with Normalize LOESS")
boxplot(t(cut_scale),
        main = "Samples with Normalized Scale")
```

``` {r}
# LOGGED CUT 1
# normalize the data
log_quantile <- normalizeBetweenArrays(log_cut, method = "quantile")
log_loess <- normalizeBetweenArrays(log_cut, method = "cyclicloess")
log_scale <- normalizeBetweenArrays(log_cut, method = "scale")

# transform to data frame for easy plotting
non_norm_log <- as.data.frame(log_cut)
log_quantile <- as.data.frame(log_quantile)
log_loess <- as.data.frame(log_loess)
log_scale <- as.data.frame(log_scale)

# look at the PROBESETS
par(mfrow=c(2,2))
boxplot(non_norm_log,
        main = "Logged, but not normalized")
boxplot(log_quantile,
        main = "Log Normalized Quantile")
boxplot(log_loess,
        main = "Log Normalized LOESS")
boxplot(log_scale,
        main = "Log Normalized Scale")

# look at the SAMPLES
par(mfrow=c(2,2))
boxplot(t(non_norm_log),
        main = "Log Samples without Norm")
boxplot(t(log_quantile),
          main = "Log Samples with Normalize Quantile")
boxplot(t(log_loess),
        main = "Log Samples with Normalize LOESS")
boxplot(t(log_scale),
        main = "Log Samples with Normalized Scale")
```
 
# GCRMA DATA!!!!! 

``` {r}
par(mfrow=c(1,2))

# look at GENES
boxplot(norm_subset,
        main = "GCRMA Normalized Probes")
boxplot(genes,
        main = "Figshare Genes")

par(mfrow=c(1,2))
# look at SAMPLES
boxplot(t(norm_subset),
        main = "GCRMA Normalized Samples")
boxplot(t(genes),
        main = "Figshare samples")
```

# GCRMA without No-UPP:

``` {r}
par(mfrow=c(1,2))

# look at GENES
boxplot(norm_sub,
        main = "GCRMA Normalized Probes, without No UPP")
boxplot(genes,
        main = "Figshare Genes")

par(mfrow=c(1,2))
# look at SAMPLES
boxplot(t(norm_sub),
        main = "GCRMA Normalized Samples without No UPP")
boxplot(t(genes),
        main = "Figshare samples")
``` 

\newpage

### Working with Figshare/gene data

We want to look at the data from Figshare to see what their "Normalized" data looks like. This way, we can compare and get a standard for what our normalized data should approximately look like.

This is what the data looks like for Figshare after they normalized. The distribution for genes are different, but the samples' distributions are similar. 

``` {r}
# look at genes
boxplot(genes,
        main = "Figshare genes")

# look at samples
boxplot(t(genes),
        main = "Figshare samples")
```


